Phylotyper: {\it In silico} predictor of subtypes from gene subtypes

### 2017-05-29 manuscript/response_to_reviewers.txt

Reviewer: 2

Comments to the Author
General comments:
Whiteside et al present here a novel python package for the prediction of bacterial molecular subtypes that are widely used in public health surveillance and outbreak analysis. The release of such a bioinformatic tools is highly valuable and timely with the democratization of WGS in microbiology laboratories worldwide. The article would benefit from some rewriting to provide more details to the reader regarding the use of Phylotyper and how to extend Phylotyper use to other bacteria than Escherichia coli.  Furthermore, a comparison of Phylotyper performance with other existing tools that the authors mention in the introduction would be welcome to assess the strength and weaknesses of this interesting phylogenetic approach.

Major comments:
Introduction:
1.      The authors should elaborate on the "current subtype systems" for readers that would not be familiar with clinical microbiology standards.

*Response*:
We added a high-level description of mechanisms currently in use to perform subtyping, namely phenotype- and DNA-based methods, with specific examples in each category. We feel this will give the readers an understanding of the breadth of subtyping methods. We also describe the characteristics of the subtypes that have been adapted for in silico prediction using WGS data. Our intention was to convey the wide range of subtyping approaches, and how certain subtyping approaches can be recapitulated from WGS data.

@@

3.      Please provide more details regarding the “assumptions of the Phylotyper approach” and how/why they would be violated when building new subtype schemes.

*Response*: 
The Phylotyper assumptions are now outlined in the Implementation section. The revised Implementation section covers the basis of the assumptions, how they can be violated, and two built-in mechanisms we developed to alert users of possible violations.

@@

Results:
4.      Whiteside et al compare here their tool with another BLAST based approach that they have developed for the comparison, and not methods published by other groups. The description of the method used for both the BLAST method and the cross validation is imprecise and would benefit from more being more detailed within the main article rather than in the supplementary material. Moreover, a comparison - at least theoretical - with other existing tools would strengthen the uniqueness of Phylotyper highlighted in the discussion.

*Response*:
Our main goal in the performance assessment was to compare how the prediction methods generalize to an independent dataset. There may exists differences in the reference training sets that underpin the tools, and these differences may mask the true performance of the method and we wanted to evalute the predictive capability of the methods themselves. SeroTypeFinder is provided as a web service. The software for SeroTypeFinder has not been made publicly available. Without access to the software, we cannot manipulate the training and test datasets to perform cross-validation analysis. The EcOH tool works directly on unassembled small reads and not assembled genome sequences. A direct comparison of the performance with these tools with Phylotyper is not possible. The intention of using a general BLAST-based approach is demonstrate the benefit of Hidden State Prediction with a phylogenetic tree over approaches that directly match input sequences against a reference database. BLAST, as used in our assessment with a percent identity and coverage cutoff, is the basis of the SeroTypeFinder tool. EcOH employs a more sophisticated algorithm for matching alleles in order to filter out spurious small reads that might align by chance. At its core, it is still matching alleles and does not develop a model of allele sequence evolution to define boundaries on subtype sequences. We feel that our performance comparison to the generalized BLAST-based method does make a valid case for using HSP in subtype prediction. We did add a direct comparison of Phylotyper to Vtyper for Stx genes. The performance results are summerized in the Results section and details on how we selected the test set are provided in Supplementary Information.

To address the reviwers comment that the BLAST and cross-validation method descriptions are imprecise, we have added more thorough descriptions of the BLAST method and the cross-validation analysis. Our aim with the revised descriptions is to provide readers sufficient information give them a complete understanding of the mechanics of the methods. Implementation details, such as percent identity cutoff or number of iterations, remain in Supplementary Information. This was done to fit with page constraints of the article type.

As suggested, we also added a theoretical comparison of the tools to the Discussion; the theoretical comparison looks at use cases, potential issues, and additional benefits outside of performance for the tools.

Discussion:
5.      There is no data to support the claim that Phylotyper easily identifies novel subtype. Did the author try to remove all alleles of one subtype and compute the predictions to see if/how phylotyper creates a consistent subtype? Also, when a new subtype is identified, does Phylotyper includes it in his typing scheme automatically? What would be the recommended tests/further analysis recommended to researchers using Phylotyper in the case a new subtype is identified?

*Response*:

6.      “Subtypes are mainly used [..] the main uses of subtype information”. Please revise this paragraph, to allow a better logical flow. Also, Debroy et al have shown the discrepancy between between serotype-based typing and the genetic similarity of O antigen gene clusters, suggesting to replace serotyping by molecular typing. They did not particularly emphasize discrepancies between existing molecular typing scheme and whole genome data as suggested by Whiteside et al.

*Response*:

We have revised this paragraph and clarified our point on the suitability of Phylotyper for predicting subtypes from WGS; namely that Phylotyper helps identify discrepancies between a subtype and genome region that is used for prediction.

@@

4.      Replace aslo by also

*Response*:
Fixed.

5.      Revise the sentence “This type of approach, there is…”.

*Response*:
Sentence replaced.

Figure 1.
6.      Please check the label that is cut on the side of the figure. Also, what does the letter at the end of each label correspond to - the expected subtype whereas the predicted subtype is color coded? Please better describe in the legend. Or remove if this is redundant information.

*Response*:
Removed the subtype letters at the end of the label; this was redundant information that was also conveyed by the color. Added information in the figure description on the meaning of the colors. Fixed labels so they fit in the window.

7.      The colors used by Phylotyper could be improved to increase readability of the output for the allele labels. There are several good packages to select colors within R or python.

*Response*:
Switched to discriminatory palette "Set1" in R's RColorBrewer package. Phylotyper uses RColorRamp and this palette to generate sufficient colors for all subtypes.

Sup. Figure 1.
8.      Please plot the legend so that it does not overall the labels

*Response*:
Change made.

9.      Labels are too close and therefore difficult to read if there were several query allele as an input.

*Response*:
Spacing increased.

@@

Supplementary table S2:
1.      Four or five E. coli schemes?

*Response*:
The E. coli serotypes available in Phylotyper comprise two schemes: O- and H-type (here incorrectly counted as one), resulting in five tested schemes. The reference to four has been removed.

### 2017-05-29 manuscript/phylotyper_revised.tex
ADDED: responses to reviewer 2.

\begin{figure}[!tpb]%figure1
\centerline{\includegraphics{fig01.eps}}
\caption{Phylogenetic tree for select Stx2 genes. 
The subtype marginal likelihood is displayed at each node as a pie chart.
The full Stx2 tree is displayed in Supplementary Figure S1.}\label{fig:01}
\end{figure}

@@

Currently, subtype schemes for \emph{Escherichia coli} (\textit{E. coli}) are available in the Phylotyper package, and in addition to Stx include intimin and serotype O- and H-types  (Supplementary Table~1).
However, the Phylotyper software also has the capability to add new subtype schemes (instructions are provided here: \url{https://github.com/superphy/insilico-subtyping}). 
Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
To add a new subtype scheme for use in Phylotyper, users require a training set of homologous sequences with assigned subtype whose phylogenetic grouping is predictive of the subtype.
Phylotyper assumes that the provided training sequences are 1) homologous, specifically, they are suitable for alignment and phylogenetic reconstruction 2) the sequence phylogeny is correlated with the subtype distribution and 3) the set is representative of the range of sequences that make up a subtype.
Where possible, checks are built-in to the \texttt{new} pipeline to ensure the assumptions of the Phylotyper approach are not violated.
Specifically, each new subtype is subject to two tests in the \texttt{new} pipeline and results are reported to the user.
The first test checks that the distribution of inter-patristic phylogenetic distances between the instances of the same subtype is both smaller and distinct from non-identical subtypes.
This test can identify isolated cases of potentially mislabelled subtype genes that are tightly clustered with other subtypes.
A second check computes the accuracy measure, F-score, through a leave-one-out cross-validation; a procedure which uses each sequence in the training set as a test input to estimate positive and negative prediction rates (Supplementary Methods).
The performance metrics: recall and F-score rapidly decrease as the correlation between the training set phylogeny and subtype distribution decreases (Supplementary Figure)
These checks will verify that the training set sequences' phylogenetic grouping is predictive of the subtype.
There is no check that can confirm the training set covers all subtypes in a particular scheme; it is important that the user monitor and update schemes as gaps are identified.
Phylotyper is designed to return a \texttt{non-significant/undetermined} result when encountering a unknown subtype that has no representative in the training set.
Phylotyper is comparatively more robust than other methods as it generates fewer Type-I errors when encountering novel alleles or subtypes not present in the training set (see Results section for details).

@@

Phylotyper marks a progression from the sequence-similarity approach widely used in \it{in silico} subtyping strategies.
It's performance extends beyond allele matching, to perform accurate inference of subtypes for input gene sequences that not in the training data or reference database.
To compare Phylotyper to a sequence similarity-based approach, we ran two validations that looked at how both methods perform when confronted with 1) gene sequence or 2) subtype class not present in the training set.
The first validation was a leave-one-out cross-validation test that iterated through each gene in the training set, retraining the prediction tools on a training set which excludes the selected test gene, and then confirming if the retrained predictor could recover the subtype of the test gene.
This validation tests how the predictors perform when run on a distinct sequence that is not in the training set.
The second validation examined how the predictors respond when presented with a gene that has a subtype not found in the reference set.
In this validation, each subtype was iterated over and all genes that are assigned the subtype were removed from the training set.
In each iteration, we recorded the number of false-positive subtype assignments when the test sequences were used as input.
The correct response for the predictors was to return a negative result, since the subtype does not exist in the training set.
For these assessments, we developed a sequence-similarity based tool that assigns putative subtypes using BLAST.
Details for the BLAST tool and how the assessment was conducted are available in supplementary methods section.
The assessment examined the five subtype schemes available in Phylotyper: Stx1, Stx2, Eae, H-type (FliC), O-type (Wzy \& Wzx).
When tasked with assigning a novel gene sequence (i.e. one that is not in the training set) in the leave-one-out validation, Phylotyper consistently had higher precision than a top-BLAST hit approach.
The average precision in Phylotyper was 0.99 versus 0.96 in the top-BLAST hit approach(result details are available in Supplementary Table S2).
In order to achieve this level of precision, the BLAST approach often had lower recall rates; it had an average recall of 0.81 compared to 0.90 with Phylotyper
Similarily, when entire subtype classes were withheld from the training set, Phylotyper had consistenly lower false positive rates for all subtypes schemes tested; the average false positive rate in this test case was 0.11, while in the BLAST approach, the average false positive rate was 0.30.
A separate assessment for the V-typer tool; a Stx subtype predictor, was run using selected Stx gene sequences from the experimentally-verified Phylotyper training set.
The test Stx genes had sufficent surrounding DNA sequence to support \it{in silico} PCR.
In total, 24 Stx gene sequences were tested with the V-typer tool and V-typer returned results for 7, all correct.
Phylotyper can correctly predict the subtype for all these genes.
Based on this level of recall, it appears conditions in the Stx subtype setting are challenging for simulated PCR.

All new or updated subtype schemes added in Phylotyper are subject to a leave-one-out cross-validation test.
The test is part of the \texttt{add} pipeline and is used to estimate the F-score of the subtype scheme. 
The F-score reflects that predictive capability of the subtype scheme. 
If the associated phylogeny for training set gene sequences are not correlated with the subtype distribution, this will be reflected in the F-score. 
To demonstrate this property, we randomly assigned subtypes for increasing proportions of the genes in the training set and computed the F-score with the leave-one-out validation for each proportion level.
The F-score and recall rapidly decrease as the proportion of altered subtypes increased (Supplementary Figure S2).

>>>>

\begin{figure}[!tpb]%figure1
\centerline{\includegraphics{fig01.eps}}
\caption{Phylogenetic tree for select Stx2 genes. 
The subtype marginal likelihood is displayed at each node as a pie chart. Subtype is indicated by color as shown in the legend. The full Stx2 tree is displayed in Supplementary Figure S1.}\label{fig:01}
\end{figure}

@@

Currently, the Phylotyper package includes the following subtype schemes for \emph{Escherichia coli} (\textit{E. coli}): Stx, intimin and serotype O- and H-types  (Supplementary Table~1).
However, the Phylotyper software also has the capability to add new subtype schemes (instructions are provided here: \url{https://github.com/superphy/insilico-subtyping}). 
Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
To add a new subtype scheme for use in Phylotyper, users require a training set of homologous sequences with assigned subtype whose phylogenetic grouping is predictive of the subtype.
Phylotyper assumes that the provided training sequences are 1) homologous, specifically, they are suitable for alignment and phylogenetic reconstruction 2) the sequence phylogeny is correlated with the subtype distribution and 3) the set is representative of the range of sequences that make up a subtype.
Where possible, checks are built-in to the \texttt{new} pipeline to validate the submitted reference set.
Specifically, each new subtype is subject to two tests in the \texttt{new} pipeline and results are reported to the user.
The first test checks that the distribution of inter-patristic phylogenetic distances between the instances of the same subtype is both smaller and distinct from non-identical subtypes.
This test can identify isolated cases of potentially mislabelled subtype genes that are tightly clustered with other subtypes.
A second check computes the accuracy measure, F-score, through a leave-one-out cross-validation; a procedure which uses each sequence in the training set as a test input to estimate positive and negative prediction rates (Supplementary Methods).
The performance metrics: recall and F-score rapidly decrease as the correlation between the training set phylogeny and subtype distribution decreases (Supplementary Figure)
These checks will verify that the training set sequences' phylogenetic grouping is predictive of the subtype.
There is no check that can confirm the training set covers all subtypes in a particular scheme; it is important that the user monitor and update schemes as gaps are identified.
Phylotyper is designed to return a \texttt{non-significant/undetermined} result when encountering a unknown subtype that has no representative in the training set.
Phylotyper is comparatively more robust than other methods as it generates fewer Type-I errors when encountering novel alleles or subtypes not present in the training set (see Results section for details).

@@

Phylotyper marks a progression from the sequence-similarity approach that is the basis of multiple \it{in silico} subtyping strategies.
It's performance extends beyond allele matching, to perform accurate inference of subtypes for input gene sequences that are not in the training data or reference database.
To compare Phylotyper to a sequence similarity-based approach, we ran two validations that looked at how both methods perform when confronted with 1) gene sequence or 2) subtype class not present in the training set.
The first validation was a leave-one-out cross-validation test that iterated through each gene in the training set, retraining the prediction tools on a training set which excludes the selected test gene, and then confirming if the retrained predictor could recover the subtype of the test gene.
This validation tests how the predictors perform when run on a distinct sequence that is not in the training set.
The second validation examined how the predictors respond when presented with a gene that has a subtype not found in the reference set.
In this validation, each subtype was iterated over and all genes that are assigned the subtype were removed from the training set.
In each iteration, we recorded the number of false-positive subtype assignments when the test sequences were used as input.
The correct response for the predictors was to return a negative result, since the subtype does not exist in the training set.
For these assessments, we developed a sequence-similarity based tool that assigns putative subtypes using BLAST.
This generalized BLAST tool assigns a subtype when the top BLAST match for an unknown query gene against an annotated reference database is above a pre-selected percent identify and alignment coverage; the query is given the subtype from the top match in this case
Details for the BLAST tool and how the assessment was conducted are available in supplementary methods section.
The assessment examined the five subtype schemes available in Phylotyper: Stx1, Stx2, Eae, H-type (FliC), O-type (Wzy \& Wzx).
When tasked with assigning a novel gene sequence (i.e. one that is not in the training set) in the leave-one-out validation, Phylotyper consistently had higher precision than a top-BLAST hit approach.
The average precision in Phylotyper was 0.99 versus 0.96 in the top-BLAST hit approach(result details are available in Supplementary Table S2).
In order to achieve this level of precision, the BLAST approach often had lower recall rates; it had an average recall of 0.81 compared to 0.90 with Phylotyper
Similarily, when entire subtype classes were withheld from the training set, Phylotyper had consistenly lower false positive rates for all subtypes schemes tested; the average false positive rate in this test case was 0.11, while in the BLAST approach, the average false positive rate was 0.30.
A separate assessment for the V-typer tool; a Stx subtype predictor, was run using selected Stx gene sequences from the experimentally-verified Phylotyper training set.
The test Stx genes had sufficent surrounding DNA sequence to support \it{in silico} PCR.
In total, 24 Stx gene sequences were tested with the V-typer tool and V-typer returned results for 7, all correct.
Phylotyper can correctly predict the subtype for all these genes.
Based on this level of recall, it appears conditions in the Stx subtype setting are challenging for simulated PCR.

All new or updated subtype schemes added in Phylotyper are subject to a leave-one-out cross-validation test.
The test is part of the \texttt{add} pipeline and is used to estimate the F-score of the subtype scheme. 
The F-score reflects that predictive capability of the subtype scheme. 
If the associated phylogeny for training set gene sequences are not correlated with the subtype distribution, this will be reflected in the F-score. 
To demonstrate this property, we randomly assigned subtypes for increasing proportions of the genes in the training set and computed the F-score with the leave-one-out validation for each proportion level.
The F-score and recall rapidly decrease as the proportion of randomly altered subtypes increases (Supplementary Figure S2).
